---
title: "Turkey in Comparison: GDP, Education and Inequality Against Neighbours and Europe (1990-2020)"
author: "Omer GIZBILI & Kaan Mert SEVINC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7
)
```

```{r libraries}
# Core libraries
library(tidyverse)  # data manipulation and visualization
library(knitr)      # report generation
library(kableExtra) # formatted tables
library(scales)     # axis formatting
library(zoo)        # linear interpolation for missing values
```

---

# Introduction

## Context

This project analyzes Turkey's economic and social development between 1990 and 2020,
comparing it to its direct neighbours and five major European economies. We focus on
three key indicators: GDP per capita, mean years of schooling, and the Gini coefficient
(income inequality).

**Research question:**
*How does Turkey's economic development and social indicators compare to its neighbours
and major European economies between 1990 and 2020?*

## Data Sources

We use three datasets from **Our World in Data**, all sourced from major international institutions:

| Dataset | Source | Period | Coverage |
|---|---|---|---|
| GDP per capita (PPP, 2021 USD) | World Bank | 1990–2024 | 234 countries |
| Mean years of schooling | Lee & Lee (2016) | 1870–2020 | 188 countries |
| Gini coefficient | Luxembourg Income Study | 1967–2021 | 169 countries |

## Description of Variables

- **GDP per capita**: Average economic output per person, adjusted for inflation and purchasing power parity (PPP). Expressed in constant 2021 international dollars.
- **Mean years of schooling**: Average number of years of education received by people aged 25 and older.
- **Gini coefficient**: Measures income inequality on a scale from 0 (perfect equality) to 1 (maximum inequality).

---

# Methodology

## Data Loading

```{r load-data}
# Load the three raw CSV files
# Paths are relative to the rapport/ folder
gdp_raw  <- read_csv("../data/raw/gdp-per-capita-worldbank.csv")
edu_raw  <- read_csv("../data/raw/mean-years-of-schooling-long-run.csv")
gini_raw <- read_csv("../data/raw/economic-inequality-gini-index.csv")
```

## Data Characterization

```{r characterization}
# Overview of each dataset: dimensions, countries, time span
datasets <- list(
  "GDP per capita"       = gdp_raw,
  "Years of schooling"   = edu_raw,
  "Gini coefficient"     = gini_raw
)

carac <- map_dfr(datasets, function(df) {
  tibble(
    Rows      = nrow(df),
    Columns   = ncol(df),
    Countries = n_distinct(df$Entity),
    Year_min  = min(df$Year, na.rm = TRUE),
    Year_max  = max(df$Year, na.rm = TRUE)
  )
}, .id = "Dataset")

kable(carac, caption = "Characterization of the three datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Data Quality Assessment

```{r missing-global}
# Missing values per dataset
missing_report <- map_dfr(datasets, function(df) {
  df %>%
    summarise(across(everything(), ~ sum(is.na(.)))) %>%
    pivot_longer(everything(),
                 names_to  = "Variable",
                 values_to = "Missing_values")
}, .id = "Dataset")

kable(missing_report, caption = "Missing values per variable") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Data Cleaning and Integration

### Step 1 — Build the complete dataset (left_join)

```{r complete-data}
# We join the three datasets on Entity and Year
# left_join keeps all GDP rows and adds schooling/gini where available
# This preserves all country-year combinations from the GDP dataset
complete_data <- gdp_raw %>%
  rename(gdp = `GDP per capita`) %>%
  select(Entity, Year, gdp) %>%
  left_join(
    edu_raw %>%
      rename(schooling = `Average years of schooling`) %>%
      select(Entity, Year, schooling),
    by = c("Entity", "Year")
  ) %>%
  left_join(
    gini_raw %>%
      rename(gini = `Gini coefficient`) %>%
      select(Entity, Year, gini),
    by = c("Entity", "Year")
  )

cat("Complete dataset dimensions:", nrow(complete_data), "rows x", ncol(complete_data), "columns\n")
cat("Countries:", n_distinct(complete_data$Entity), "\n")
cat("Years:", min(complete_data$Year), "to", max(complete_data$Year), "\n")
```

### Step 2 — Define the country list

```{r countries}
# Turkey + 8 neighbours + 5 major European economies
countries <- c(
  "Turkey",
  # Neighbours
  "Greece", "Bulgaria", "Georgia", "Armenia", "Azerbaijan",
  "Iran", "Iraq", "Syria",
  # Major European economies
  "Spain", "United Kingdom", "France", "Germany", "Italy"
)

cat("Total countries selected:", length(countries), "\n")
```

### Step 3 — Filter the analysis dataset

```{r analysis-data}
# Keep only selected countries and years 1990-2020
analysis_data <- complete_data %>%
  filter(Entity %in% countries,
         Year >= 1990, Year <= 2020)

cat("Analysis dataset:", nrow(analysis_data), "rows\n")
```

### Step 4 — Data coverage per country

```{r coverage}
# How many non-missing observations per country per variable?
coverage <- analysis_data %>%
  group_by(Entity) %>%
  summarise(
    n_gdp       = sum(!is.na(gdp)),
    n_schooling = sum(!is.na(schooling)),
    n_gini      = sum(!is.na(gini))
  ) %>%
  arrange(desc(n_gdp))

kable(coverage,
      caption = "Data coverage per country (1990-2020)",
      col.names = c("Country", "GDP observations", "Schooling observations", "Gini observations")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Observations:**

- GDP data is complete for all 14 countries (31 observations each).
- Schooling data is available roughly every 5 years; Azerbaijan and Georgia have no data.
- Gini data is highly incomplete for Iraq (2 obs.), Syria (4 obs.), and Azerbaijan (6 obs.).

---

# Analysis

## GDP per Capita

### All countries

```{r plot-all-gdp}
# All 14 countries, linetype distinguishes groups
# Turkey is highlighted with a solid line
analysis_data %>%
  mutate(group = ifelse(Entity == "Turkey", "Turkey",
                 ifelse(Entity %in% c("Greece", "Bulgaria", "Georgia",
                                       "Armenia", "Azerbaijan", "Iran",
                                       "Iraq", "Syria"),
                        "Neighbour", "Europe"))) %>%
  ggplot(aes(x = Year, y = gdp, color = Entity, linetype = group)) +
  geom_line(linewidth = 1) +
  geom_point(data = . %>% filter(Year %in% c(1990, 1995, 2000, 2005, 2010, 2015, 2020)),
             size = 2) +
  scale_y_continuous(labels = label_dollar(big.mark = ","),
                     breaks = seq(0, 60000, by = 5000)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  scale_linetype_manual(values = c("Turkey" = "solid",
                                   "Neighbour" = "dashed",
                                   "Europe" = "dotted")) +
  labs(title   = "GDP per capita — Turkey vs Neighbours & Europe (1990-2020)",
       x       = "Year", y = "GDP per capita (USD)",
       color   = "Country", linetype = "Group",
       caption = "Source: World Bank via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Neighbours

```{r plot-neighbours-gdp}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Greece", "Bulgaria", "Georgia",
                        "Armenia", "Azerbaijan", "Iran", "Iraq", "Syria")) %>%
  ggplot(aes(x = Year, y = gdp, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(data = . %>% filter(Year %in% c(1990, 1995, 2000, 2005, 2010, 2015, 2020)),
             size = 3) +
  scale_y_continuous(labels = label_dollar(big.mark = ","),
                     breaks = seq(0, 40000, by = 5000)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "GDP per capita — Turkey vs Neighbours (1990-2020)",
       x       = "Year", y = "GDP per capita (USD)", color = "Country",
       caption = "Source: World Bank via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Europe

```{r plot-europe-gdp}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Spain", "United Kingdom",
                        "France", "Germany", "Italy")) %>%
  ggplot(aes(x = Year, y = gdp, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(data = . %>% filter(Year %in% c(1990, 1995, 2000, 2005, 2010, 2015, 2020)),
             size = 3) +
  scale_y_continuous(labels = label_dollar(big.mark = ","),
                     breaks = seq(0, 60000, by = 5000)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "GDP per capita — Turkey vs Europe (1990-2020)",
       x       = "Year", y = "GDP per capita (USD)", color = "Country",
       caption = "Source: World Bank via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

## Mean Years of Schooling

### All countries

```{r plot-all-schooling}
analysis_data %>%
  filter(!is.na(schooling)) %>%
  mutate(group = ifelse(Entity == "Turkey", "Turkey",
                 ifelse(Entity %in% c("Greece", "Bulgaria", "Georgia",
                                       "Armenia", "Azerbaijan", "Iran",
                                       "Iraq", "Syria"),
                        "Neighbour", "Europe"))) %>%
  ggplot(aes(x = Year, y = schooling, color = Entity, linetype = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(breaks = seq(0, 15, by = 1)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  scale_linetype_manual(values = c("Turkey" = "solid",
                                   "Neighbour" = "dashed",
                                   "Europe" = "dotted")) +
  labs(title   = "Mean years of schooling — Turkey vs Neighbours & Europe (1990-2020)",
       x       = "Year", y = "Mean years of schooling",
       color   = "Country", linetype = "Group",
       caption = "Source: Lee & Lee (2016) via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Neighbours

```{r plot-neighbours-schooling}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Greece", "Bulgaria", "Georgia",
                        "Armenia", "Azerbaijan", "Iran", "Iraq", "Syria"),
         !is.na(schooling)) %>%
  ggplot(aes(x = Year, y = schooling, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_y_continuous(breaks = seq(0, 15, by = 1)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "Mean years of schooling — Turkey vs Neighbours (1990-2020)",
       x       = "Year", y = "Mean years of schooling", color = "Country",
       caption = "Source: Lee & Lee (2016) via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Europe

```{r plot-europe-schooling}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Spain", "United Kingdom",
                        "France", "Germany", "Italy"),
         !is.na(schooling)) %>%
  ggplot(aes(x = Year, y = schooling, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_y_continuous(breaks = seq(0, 15, by = 1)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "Mean years of schooling — Turkey vs Europe (1990-2020)",
       x       = "Year", y = "Mean years of schooling", color = "Country",
       caption = "Source: Lee & Lee (2016) via Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

## Gini Coefficient

Missing values are interpolated linearly using `zoo::na.approx()` to allow continuous
line plotting. Original missing values are preserved in `analysis_data`.

### All countries

```{r plot-all-gini}
analysis_data %>%
  mutate(group = ifelse(Entity == "Turkey", "Turkey",
                 ifelse(Entity %in% c("Greece", "Bulgaria", "Georgia",
                                       "Armenia", "Azerbaijan", "Iran",
                                       "Iraq", "Syria"),
                        "Neighbour", "Europe"))) %>%
  group_by(Entity) %>%
  mutate(gini = zoo::na.approx(gini, na.rm = FALSE)) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = gini, color = Entity, linetype = group)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(data = . %>% filter(!is.na(gini)), size = 2) +
  scale_y_continuous(breaks = seq(0, 0.6, by = 0.05)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  scale_linetype_manual(values = c("Turkey" = "solid",
                                   "Neighbour" = "dashed",
                                   "Europe" = "dotted")) +
  labs(title   = "Gini coefficient — Turkey vs Neighbours & Europe (1990-2020)",
       x       = "Year", y = "Gini coefficient",
       color   = "Country", linetype = "Group",
       caption = "Source: Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Neighbours

```{r plot-neighbours-gini}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Greece", "Bulgaria", "Georgia",
                        "Armenia", "Azerbaijan", "Iran", "Iraq", "Syria")) %>%
  group_by(Entity) %>%
  mutate(gini = zoo::na.approx(gini, na.rm = FALSE)) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = gini, color = Entity)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(data = . %>% filter(!is.na(gini)), size = 3) +
  scale_y_continuous(breaks = seq(0, 0.6, by = 0.05)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "Gini coefficient — Turkey vs Neighbours (1990-2020)",
       x       = "Year", y = "Gini coefficient", color = "Country",
       caption = "Source: Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

### Turkey vs Europe

```{r plot-europe-gini}
analysis_data %>%
  filter(Entity %in% c("Turkey", "Spain", "United Kingdom",
                        "France", "Germany", "Italy")) %>%
  group_by(Entity) %>%
  mutate(gini = zoo::na.approx(gini, na.rm = FALSE)) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = gini, color = Entity)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(data = . %>% filter(!is.na(gini)), size = 3) +
  scale_y_continuous(breaks = seq(0, 0.6, by = 0.05)) +
  scale_x_continuous(breaks = seq(1990, 2020, by = 5)) +
  labs(title   = "Gini coefficient — Turkey vs Europe (1990-2020)",
       x       = "Year", y = "Gini coefficient", color = "Country",
       caption = "Source: Our World in Data") +
  theme_minimal(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

## Correlation Analysis

```{r correlation}
# Overall correlations across all 14 countries
cat("=== Overall correlations ===\n")
cat("GDP ~ Schooling:", round(cor(analysis_data$gdp, analysis_data$schooling, use = "complete.obs"), 3), "\n")
cat("GDP ~ Gini:     ", round(cor(analysis_data$gdp, analysis_data$gini,      use = "complete.obs"), 3), "\n\n")

# Per-country correlations (only for countries with at least 3 data points)
cor_table <- analysis_data %>%
  group_by(Entity) %>%
  summarise(
    cor_gdp_schooling = ifelse(sum(!is.na(schooling)) >= 3,
                               round(cor(gdp, schooling, use = "complete.obs"), 3), NA),
    cor_gdp_gini      = ifelse(sum(!is.na(gini)) >= 3,
                               round(cor(gdp, gini, use = "complete.obs"), 3), NA)
  )

kable(cor_table,
      caption = "Pearson correlations per country",
      col.names = c("Country", "GDP ~ Schooling", "GDP ~ Gini")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(which(cor_table$Entity == "Turkey"), bold = TRUE, background = "#fff3f3")
```

**Key observations:**

- Turkey shows a very strong positive correlation between GDP and schooling (r = 0.972), similar to France (0.973), Germany (0.963) and Iran (0.967).
- The GDP ~ Gini correlation for Turkey (0.449) suggests that economic growth was accompanied by a moderate increase in inequality.
- Syria shows a negative GDP ~ Schooling correlation (-0.263), likely due to the collapse of GDP during the civil war while education metrics remained relatively stable.

## Regression Analysis

### Model 1 — GDP ~ Schooling

```{r model-schooling}
# Simple linear regression: does education predict GDP?
model_schooling <- lm(gdp ~ schooling, data = analysis_data)
summary(model_schooling)
```

**Interpretation:** Each additional year of schooling is associated with an average increase
of **$5,780** in GDP per capita. The model explains **60.4%** of the variance in GDP (R² = 0.604).
The p-value is extremely small (< 2.2e-16), confirming the statistical significance.

### Model 2 — GDP ~ Gini

```{r model-gini}
# Simple linear regression: does inequality predict GDP?
model_gini <- lm(gdp ~ gini, data = analysis_data)
summary(model_gini)
```

**Interpretation:** The Gini coefficient alone explains only **15.9%** of GDP variance (R² = 0.159).
The negative coefficient (-177,033) suggests that higher inequality is associated with lower GDP,
but the model's explanatory power is weak.

### Model 3 — GDP ~ Schooling + Gini

```{r model-full}
# Multiple regression: both variables together
model_full <- lm(gdp ~ schooling + gini, data = analysis_data)
summary(model_full)
```

**Interpretation:** When combining both variables, R² drops to **0.440** compared to Model 1.
This is because many observations with Gini data are missing, reducing the sample size from
79 to 51 observations. Schooling remains highly significant (p < 0.001) while Gini is not
significant (p = 0.116) when controlling for schooling.

---

# Conclusion

This analysis compared Turkey's GDP per capita, educational attainment, and income inequality
against its neighbours and five major European economies over the period 1990–2020.

**Main findings:**

- Turkey's GDP per capita grew significantly over the period, surpassing most of its neighbours
  but remaining well below the major European economies.
- Turkey's mean years of schooling increased steadily, reflecting sustained public investment
  in education. By 2020, Turkey's educational level was comparable to Iran and Bulgaria,
  but still below Greece and the European economies.
- Turkey's Gini coefficient remained relatively high (around 0.40–0.43) compared to European
  countries, indicating persistent income inequality despite economic growth.
- Regression analysis confirms that education is the strongest predictor of GDP per capita
  in this sample (R² = 0.604, p < 0.001), with each additional year of schooling associated
  with approximately $5,780 in additional GDP per capita.

**Limitations:**

- Schooling data is only available every 5 years; linear interpolation was used for visualization.
- Gini data is highly incomplete for several countries (Azerbaijan, Iraq, Syria).
- Correlation and regression do not imply causality.
- Country-level GDP per capita masks internal regional disparities.

**Future directions:**

- Incorporate regional data (e.g. NUTS regions for Turkey and EU countries).
- Analyse gender gaps in education and their relation to inequality.
- Extend the time period to 2024 using the available GDP data.

---

# References

- Eurostat, OECD, IMF and World Bank (2026). *GDP per capita – World Bank – In constant international-$*. Our World in Data. https://ourworldindata.org/grapher/gdp-per-capita-worldbank

- Lee, J.-W. and Lee, H. (2016). Human capital in the long run. *Journal of Development Economics*, 122, 147–169.

- R Core Team (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria.

- Wickham, H. et al. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, 4(43), 1686.

- Zeileis, A. and Grothendieck, G. (2005). zoo: S3 infrastructure for regular and irregular time series. *Journal of Statistical Software*, 14(6), 1–27.
